Insecticide rotations delay evolution of resistance in a minority of model runs when compared to sequential use.

### 2018-05-10 inst/documents/paper3/paper3_rotations.tex
pap3

---
title: "Insecticide rotations delay evolution of resistance in a minority of model runs when compared to sequential use."
# output:
#   pdf_document:
#     keep_tex: true
#     fig_caption: true
#     latex_engine: pdflatex

output: word_document
#author: "Andy South and Ian Hastings"
#date: "`r Sys.Date()`"
bibliography: library.bib
csl: biomed-central.csl
fontsize: 11pt
spacing: double
# below worked in test_linenumbers.Rmd
header-includes:
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
---

@@

southandy@gmail.com, andy.south@lstmed.ac.uk

ian.hastings@lstmed.ac.uk

\* Corresponding author

```{r, eval=TRUE, include=FALSE}
#!!!! PAPER PROTOCOL
#!!!! submission MS, knit to word_document 
#!!!!  in WORD: delete figs, set double space, add line numbers 
#!!!!  (spacing & line numbers work for pdf output but not for word)
library(knitr)
opts_chunk$set(dev="tiff",
               dev.args=list(compression="lzw"),
               dpi=300,
               cache=FALSE,
               fig.path='pap3figs/',
               echo=FALSE, warning=FALSE, message=FALSE)

#don't work for word
#fig.show='hold', fig.align='center'
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
  library(rotations)
  library(ggplot2)
  library(gridExtra)

  outFolder <- "C:\\Dropbox\\resistanceResults\\rotations\\"

  # load simulation outputs for some plots
  # beware that each object is called df_res_all
  # some plots are created by directly running the model
  load("df_res_all_rot_nocost_nomig1000.rda") #nocost nomig  
  df_nocost_nomig <- df_res_all
  load("df_res_all_rot10000.rda") #cost
  df_cost <- df_res_all
  load("df_res_all_rot_mig10000.rda") #mig 10k
  df_mig <- df_res_all
  load("df_res_all_rot_cost_mig10000.rda") #costmig 10k
  df_cost_mig <- df_res_all

```

## Abstract

@@

## Introduction
Despite much field and modelling work on the evolution of resistance under different insecticide or drug strategies [@Consortium2013] there is a small evidence base to support decisions about the use of insecticides in public health [@Sternberg2017].

Rotations are advocated as one of the most favoured approaches for Insecticide Resistance Management (IRM)[@WHO2012].

In a rotation the insecticide used is changed on a set time interval irrespective of resistance levels. Sequential use, in contrast, denotes the practice of using an insecticide until resistance to it reaches a certain threshold and then switching to another. Different rotation time intervals can be considered, e.g. in vector control for Malaria, Indoor Residual Spraying may be repeated every year (around 10 anopheles mosquito generations) and Insecticide Treated Nets may be replaced every two years (around 20 generations). Multiples of these intervals could be used as the rotation interval. Multiple generational rotations like this contrast with the situation in agriculture where a rotation is sometimes more tightly defined as having an interval of a single generation [@Sudo2017].

@@

## Methods

A population genetic model was developed tracking the change in resistance allele frequency as consequence of the genetic determination of the resistance phenotype, and the deployment policy for the insecticides. The model uses a similar approach to that previously used to compare insecticide mixtures and sequences [@Levick2017,@South2018]. The implementation is simpler here because for rotations and sequences there is only a need to follow one insecticide at a time and linkage disequilibrium can be ignored. The simpler implementation allows the model to be run for an unlimited number of insecticides. The algebra behind the model is described in the supplementary information.

1. Sequences : one insecticide is used until resistance threshold frequency (0.5 in this case) is reached, then switch to a new insecticide.

2. Rotations : use insecticide for set time (i.e. number of insect generations), switch to another.

In both cases the choice of the new insecticide to use was made by going through the list of available insecticides and choosing the next one for which resistance was below the threshold. Thus there is the potential to switch back to insecticides that had been used earlier in a model run if resistance had not reached the threshold or if resistance had declined below the threshold during a period when not in use. This means that the “sequential” policy is effectively a “rotate when resistant” policy but we use the former term for consistency with previous work. *todo cite different terms from Rex consortium review* For both rotations and sequences the simulation continues until no more insecticides remain below their resistance thresholds.

@@

**Table 1. Model inputs and the ranges used in simulations**

Input  | Description | Min | Max
---------------------- | ------------------------------------- | ----- | -----
Effectiveness | proportion of susceptible (SS) insects killed by exposure to insecticide | 0.5 | 1
Exposure | proportion of females exposed to insecticide | 0.4 | 0.9
Male exposure | proportion of males exposed to insecticide as proportion of females | 0 | 1
Resistance restoration  | ability of resistance (RR) to restore fitness of insects exposed to insecticide | 0.1 | 0.9
Dominance of resistance | sets fitness of heterozygous (SR) insects between that of SS & RR in presence of insecticide | 
Frequency  | frequency of resistance alleles within the population | 0.005 | 0.1
Cost of resistance  |  fitness of resistant (RR) insects in absence of insecticide | 0 | 0.1
Dominance of cost  | sets fitness of heterozygous (SR) insects between that of SS & RR in absence of insecticide | 0.1 | 0.9
Insecticide number | number of independent insecticides available | 2 | 5  
Rotation interval | number of generations to use insecticide before switching | 5 | 50
Coverage | for scenarios with dispersal, the proportion of insects in the treated area rather than untreated refugia |  0.1 | 0.9
Dispersal | proportion of population exchanged between treated and untreated areas per generation, 0=none, 1=random mixing |  0.1 | 0.9

@@

1. baseline : no resistance fitness costs or dispersal to/from untreated refugia
2. fitness costs of resistance added
3. dispersal link to untreated areas (no costs)
4. costs and dispersal

@@

#### Assessing relative performance of insecticide use strategies

*todo change this bit, we could actually use time-to-resistance for all strategies ?*
*make it clearer exactly what we do*
It is necessary to choose a measure to quantify the performance of an insecticide-use strategy and enable comparison with an alternative. For this we summed the number of generations when an insecticide was deployed and the resistance allele frequency for that insecticide was below 0.5. This allowed comparison between strategies using a variable number of insecticides. Previously time-to-resistance, namely the number of generations it takes to reach a defined resistance threshold (usually a resistance allele frequency of 0.5), was used. Using time-to-resistance worked when considering just two insecticides, but when the number of insecticides is a variable it is not sufficient.

## Results

#### Results1. form of resistance evolution with and without costs and dispersal.

@@

*todo continue describing the example runs first, before talking about overall results*

### Results2 : comparing rotations and sequences across model runs.

In all the example runs shown the resistance threshold (allele frequency of 0.5) was reached for all insecticides within the course 500 generations which was the maximum limit imposed in the model. There were many model runs [*quantify later when run finalised*] in which resistance frequencies for at least some insecticides were below thresholds at the end of 500 generations. Such a run could be considered a successful strategy given the conditions of the scenario because insecticides below the resistance threshold are still available to be used.

@@

The number of model runs in each of these groups for the different scenarios are shown in Fig 5. In the base scenario with no costs or dispersal (Fig 5A) the majority of runs do not succeed, a small proportion succeed for both strategies and none succeed for either strategy alone. When costs of reistance are added (Fig 5B) more runs succeed for both strategies, up to about half of those that succeed for neither, and a small proportion (< 4%) succeed for rotations alone. Adding both costs and dispersal results in more runs succeeding for both than neither (Fig 5C).

To look at the potential benefit of rotations the number of generations below resistance thresholds was compared across strategies. This will be influenced by the previous analysis because only those runs that do not succeed for both strategies can generate a difference between them.

When there were no costs or dispersal there was no difference between rotations and sequences in the number of generations below the resistance threshold (Fig 6A) (for comparison with other strategies the results are grouped into those between plus and minus 1%). With fitness costs included 80% of model runs, similar to those in the previous section, gave no difference between rotations and sequences in the number of generations below the resistance threshold. In contrast, 7% of model runs generated an advantage of greater than 20% for rotations over sequences (Fig 6B). Dispersal from untreated refugia gives smaller benefits of rotations over sequences and in fewer runs (Fig 6C). With both costs and dispersal the benefits of rotations are similar to when runs have costs alone (Fig 6D).

#### Runs where rotations have substantial advantage over sequences

*show violin plots here ?*

@@

## Discussion

The model shows that rotations and sequences most often have very similar performance (within 1 %) in terms of time to reach resistance thresholds (Fig 6).

*todo add something about how costs and dispersal generally improve the performance of both strategies*
*todo add about how rotations generally perform better relative to sequences under 'intermediate' scenarios. In 'better' scenarios both strategies do equally well and in 'poorer' scenarios they do equally badly.*

1. rotations and sequences give the same results when there are no costs or dispersal.
2. with resistance costs, 80% of runs are the same for rotations and sequences, but 6-7% of runs have > 20% advantage for rotations
3. with dispersal from untreated refugia but no costs, small advantage of rotations but very few runs < 0.1% have >20% advantage.

This can be understood intuitively. When there are no costs, the dynamics are the same, it is simply that sequential deployment proceeds in a few, larger steps while rotations proceeds in more numerous smaller steps; both reach the same point in around the same time. However dominance effects on cost and resistance can act to keep frequencies low in a rotations policy. Recall that at low frequencies of resistance most resistance alleles are present as heterozygotes RS and only a vey small proportions heterozygotes. However costs are dominant and still act at low frequencies. This is illustrated on Figure Y. Relatively rapid rotations keep the allele frequencies at the LH side of Figure Y where costs during the periods when the insecticide is not deployed greatly outweigh the selection for IR during the periods when the insecticide is deployed. We believe the explanation is therefore that rotations keep allele frequencies in the parameter space to the left of Figure Y and hence acts to slow the evolution of IR compared to sequences which allow frequencies to rise to relatively high levels before the insecticide is replaced.  (does this sound reasonable??). Note also that starting frequency, as expected, had an impact (Figure 7), lower starting frequencies slightly favouring rotations by allowing more time in the “low frequency” parameter space.

In rare circumstances, when there are costs of resistance, rotations can lead to the resistance frequencies for a group of insecticides being kept below resistance thresholds for the length of the simulations where sequences run out of insecticides within around 100 generations. Is it possible that these rare circumstances where parameter values are 'just-right' could be implemented operationally ? The problem is that these rare circumstances rely on high costs of resistance, high dominance of these costs and low dominance of the selection for the resistance itself. These parameters are all not easy to measure and there isn't agreement on their likely values [@Kliot2012, @Bourguet2000].

Note however that this combination of recessive resistance and high dominance costs are exactly the features most desirable in any insecticide irrespective of how it is deployed. We discuss fitness below but note here that it is very difficult to maintain insecticide concentration at the level required to make resistance recessive. Concentrations decay post-application in most deployment strategies and this cause resistance to gradually change from being recessive to being dominant; see Figure 1 of [@Levick2017] and our more extensive discussion in [cite South et al WoS manuscript].

The importance of resistance costs for substantial evolutionary benefits of rotations over sequences points to the importance of the debate over the fitness costs of insecticide resistance e.g. [@Kliot2012].
*extend this discussion of fitness costs*

We also found frequency of rotations had little effect. This is important in the control of vectors of human diseases such as malaria and dengue. These predominantly occur in resource-poor areas often with poor healthcare infrastructure. Experience suggests that rotation targets would often be missed (e.g rotations occur approximately every 15 generations rather than the planned “official” 10 generations) but our simulations suggest this would not fatally undermine a rotations policy.

### Caveats

The model does not include the potential effects of modifier genes ameliorating the costs of resistance alleles. To include this would require a more detailed model tracking linkage disequilibrium between each resistance locus and its modifier. It has been suggested that rotations could be favoured as a strategy because they keep resistance frequences lower and restrict selection pressure for modifier genes [@WHO2012]. However, there is little evidence to support the existence of such modifier genes in insectide resistance []. Even with the existence of modifier genes it is unlikely that they would provide much of an advantage to rotations in the scenarios we describe. In the modelled scenarios the use of an insecticide is stopped when the resistance frequency for that insecticide reaches 0.5. Thus in all scenarios resistance frequencies do not remain at high levels for long. In this situation it seems unlikely that modifiers would be selected for.

### Conclusions

In their effect on the evolution of insecticide resistance, rotations and sequences are most often the same. If costs of resistance are included, infrequently (less than 7% of model runs set between plausible limits) there are substantial benefits to the rotation strategy over a sequence. There was never a substantial benefit of a sequence over a rotation. Whether or not insecticide rotations are predicted to offer sizeable benefits over sequences is dependent on coarse and fine scale issues about the nature of insecticide resistance that are yet to be agreed upon. The model shows that resistance costs are necessary to provide a substantial evolutionary advantage to rotations. However even when these requirements are satisfied particular combinations of inputs are required to generate the advantage of rotations.

@@

## TODO
*add to methods description of how fitness determined in model, move & modify fig 1*
*try to come up with brief name for model output to make it easier to refer to it later*
*add caveat about polygenic resistance*
*The caption to each graph needs to have the list of parameter values used to generate them*
*work on Ians fig Y to help explanation*

@@

## Figures

```{r Fig1, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7.5, fig.show='hold', fig.align='left', fig.cap="Figure 1. [todo need to modify caption from MJ paper, start by saying that only in left panel during application of the insecticide] The effect of model inputs on the fitness of genotypes for a single insecticide. Fitness is shown on the y-axis and the different genotypes (SS, SR, RR) on the x axis. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS, resistance restoration 'restores' a portion of the fitness for RR and dominance of resistance determines how the fitness for SR lies between that of SS and RR. For those that are not exposed, fitness of SS is set to 1 by definition, resistance cost determines the fitness of RR and again dominance of cost determines how the fitness for SR sits between that of SS and RR. In this example effectiveness=0.8, resistance restoration=0.5 which 'restores' half of the fitness lost due to the insecticide, dominance of resistance=0.7 which sets the fitness of the SR closer to RR than SS. Resistance cost=0.3 which reduces fitness in the absence of the insecticide from 1 to 0.7, and dominance of cost=0.8 which sets fitness of SR close to RR."}

library(resistance) 
plot_fit_calc( effectiveness=0.8, resistance_restoration=0.5, dominance_restoration=0.7, dominance_cost=0.7, cost=0.3 )

```

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Insecticide rotations delay evolution of resistance in a minority of model runs when compared to sequential use.},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Insecticide rotations delay evolution of resistance in a minority of
model runs when compared to sequential use.}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers

\begin{document}
\maketitle

@@

\href{mailto:southandy@gmail.com}{\nolinkurl{southandy@gmail.com}},
\href{mailto:andy.south@lstmed.ac.uk}{\nolinkurl{andy.south@lstmed.ac.uk}}

\href{mailto:ian.hastings@lstmed.ac.uk}{\nolinkurl{ian.hastings@lstmed.ac.uk}}

* Corresponding author

\subsection{Abstract}\label{abstract}

@@

\subsection{Introduction}\label{introduction}

Despite much field and modelling work on the evolution of resistance
under different insecticide or drug strategies {[}1{]} there is a small
evidence base to support decisions about the use of insecticides in
public health {[}2{]}.

Rotations are advocated as one of the most favoured approaches for
Insecticide Resistance Management (IRM){[}3{]}.

In a rotation the insecticide used is changed on a set time interval
irrespective of resistance levels. Sequential use, in contrast, denotes
the practice of using an insecticide until resistance to it reaches a
certain threshold and then switching to another. Different rotation time
intervals can be considered, e.g.~in vector control for Malaria, Indoor
Residual Spraying may be repeated every year (around 10 anopheles
mosquito generations) and Insecticide Treated Nets may be replaced every
two years (around 20 generations). Multiples of these intervals could be
used as the rotation interval. Multiple generational rotations like this
contrast with the situation in agriculture where a rotation is sometimes
more tightly defined as having an interval of a single generation
{[}4{]}.

@@

\subsection{Methods}\label{methods}

A population genetic model was developed tracking the change in
resistance allele frequency as consequence of the genetic determination
of the resistance phenotype, and the deployment policy for the
insecticides. The model uses a similar approach to that previously used
to compare insecticide mixtures and sequences {[}5{]}. The
implementation is simpler here because for rotations and sequences there
is only a need to follow one insecticide at a time and linkage
disequilibrium can be ignored. The simpler implementation allows the
model to be run for an unlimited number of insecticides. The algebra
behind the model is described in the supplementary information.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sequences : one insecticide is used until resistance threshold
  frequency (0.5 in this case) is reached, then switch to a new
  insecticide.
\item
  Rotations : use insecticide for set time (i.e.~number of insect
  generations), switch to another.
\end{enumerate}

In both cases the choice of the new insecticide to use was made by going
through the list of available insecticides and choosing the next one for
which resistance was below the threshold. Thus there is the potential to
switch back to insecticides that had been used earlier in a model run if
resistance had not reached the threshold or if resistance had declined
below the threshold during a period when not in use. This means that the
``sequential'' policy is effectively a ``rotate when resistant'' policy
but we use the former term for consistency with previous work.
\emph{todo cite different terms from Rex consortium review} For both
rotations and sequences the simulation continues until no more
insecticides remain below their resistance thresholds.

@@

\textbf{Table 1. Model inputs and the ranges used in simulations}

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.28\columnwidth}\raggedright\strut
Input\strut
\end{minipage} & \begin{minipage}[b]{0.46\columnwidth}\raggedright\strut
Description\strut
\end{minipage} & \begin{minipage}[b]{0.07\columnwidth}\raggedright\strut
Min\strut
\end{minipage} & \begin{minipage}[b]{0.07\columnwidth}\raggedright\strut
Max\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Effectiveness\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
proportion of susceptible (SS) insects killed by exposure to
insecticide\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.5\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
1\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Exposure\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
proportion of females exposed to insecticide\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.4\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.9\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Male exposure\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
proportion of males exposed to insecticide as proportion of
females\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
1\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Resistance restoration\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
ability of resistance (RR) to restore fitness of insects exposed to
insecticide\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.1\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.9\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Dominance of resistance\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
sets fitness of heterozygous (SR) insects between that of SS \& RR in
presence of insecticide\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Frequency\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
frequency of resistance alleles within the population\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.005\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.1\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Cost of resistance\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
fitness of resistant (RR) insects in absence of insecticide\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.1\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Dominance of cost\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
sets fitness of heterozygous (SR) insects between that of SS \& RR in
absence of insecticide\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.1\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.9\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Insecticide number\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
number of independent insecticides available\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
2\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
5\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Rotation interval\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
number of generations to use insecticide before switching\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
5\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
50\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Coverage\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
for scenarios with dispersal, the proportion of insects in the treated
area rather than untreated refugia\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.1\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.9\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright\strut
Dispersal\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
proportion of population exchanged between treated and untreated areas
per generation, 0=none, 1=random mixing\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.1\strut
\end{minipage} & \begin{minipage}[t]{0.07\columnwidth}\raggedright\strut
0.9\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  baseline : no resistance fitness costs or dispersal to/from untreated
  refugia
\item
  fitness costs of resistance added
\item
  dispersal link to untreated areas (no costs)
\item
  costs and dispersal
\end{enumerate}

@@

\paragraph{Assessing relative performance of insecticide use
strategies}\label{assessing-relative-performance-of-insecticide-use-strategies}

\emph{todo change this bit} It is necessary to choose a measure to
quantify the performance of an insecticide-use strategy and enable
comparison with an alternative. For this we summed the number of
generations when an insecticide was deployed and the resistance allele
frequency for that insecticide was below 0.5. This allowed comparison
between strategies using a variable number of insecticides. Previously
time-to-resistance, namely the number of generations it takes to reach a
defined resistance threshold (usually a resistance allele frequency of
0.5), was used. Using time-to-resistance worked when considering just
two insecticides, but when the number of insecticides is a variable it
is not sufficient.

\subsection{Results}\label{results}

\paragraph{Results1. form of resistance evolution with and without costs
and
dispersal.}\label{results1.-form-of-resistance-evolution-with-and-without-costs-and-dispersal.}

@@

\emph{todo continue describing the example runs first, before talking
about overall results}

\subsubsection{Results2 : comparing rotations and sequences across model
runs.}\label{results2-comparing-rotations-and-sequences-across-model-runs.}

In all the example runs shown the resistance threshold (allele frequency
of 0.5) was reached for all insecticides within the course 500
generations which was the maximum limit imposed in the model. There were
many model runs {[}\emph{quantify later when run finalised}{]} in which
resistance frequencies for at least some insecticides were below
thresholds at the end of 500 generations. Such a run could be considered
a successful strategy given the conditions of the scenario because
insecticides below the resistance threshold are still available to be
used.

@@

The number of model runs in each of these groups for the different
scenarios are shown in Fig 5.

\emph{todo : change the \%s because `succeeded' runs now left in} When
there were no costs or dispersal there was no difference between
rotations and sequences in the number of generations below the
resistance threshold (Fig 6A). With fitness costs included 67\% of model
runs, similar to those in the previous section, gave no difference
between rotations and sequences in the number of generations below the
resistance threshold. In contrast, 7\% of model runs generated an
advantage of greater than 20\% for rotations over sequences (Fig 6B).
Dispersal from untreated refugia gives smaller benefits of rotations
over sequences and in fewer runs (Fig 6C). With both costs and dispersal
the benefits of rotations are similar to when runs have costs alone (Fig
6D).

\paragraph{Runs where rotations have substantial advantage over
sequences}\label{runs-where-rotations-have-substantial-advantage-over-sequences}

\emph{Talk about classing runs into both `succeeded', both `failed' or
just rotation succeeded. }show violin plots here ?*

@@

\subsection{Discussion}\label{discussion}

Comparing time below resistance thresholds for rotations and sequences,
the model shows :

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  rotations and sequences the same when no costs or dispersal
\item
  with resistance costs, 67\% of runs the same for rotations and
  sequences, but 6-7\% of runs have \textgreater{} 20\% advantage for
  rotations
\item
  with dispersal from untreated refugia but no costs, small advantage of
  rotations but very few runs \textless{} 0.1\% have \textgreater{}20\%
  advantage.
\end{enumerate}

This can be understood intuitively. When there are no costs, the
dynamics are the same, it is simply that sequential deployment proceeds
in a few, larger steps while rotations proceeds in more numerous smaller
steps; both reach the same point in around the same time (e.g.~figs 1
and 3). However dominance effects on cost and resistance can act to keep
frequencies low in a rotations policy. Recall that at low frequencies of
resistance most resistance alleles are present as heterozygotes RS and
only a vey small proportions heterozygotes. However costs are dominant
and still act at low frequencies. This is illustrated on Figure Y.
Relatively rapid rotations keep the allele frequencies at the LH side of
Figure Y where costs during the periods when the insecticide is not
deployed greatly outweigh the selection for IR during the periods when
the insecticide is deployed. We believe the explanation is therefore
that rotations keep allele frequencies in the parameter space to the
left of Figure Y and hence acts to slow the evolution of IR compared to
sequences which allow frequencies to rise to relatively high levels
before the insecticide is replaced. (does this sound reasonable??). Note
also that starting frequency, as expected, had an impact (Figure 7),
lower starting frequencies slightly favouring rotations by allowing more
time in the ``low frequency'' parameter space.

In rare circumstances, when there are costs of resistance, rotations can
lead to the resistance frequencies for a group of insecticides being
kept below resistance thresholds for the length of the simulations where
sequences run out of insecticides within around 100 generations. Is it
possible that these rare circumstances where parameter values are
`just-right' could be implemented operationally ? The problem is that
these rare circumstances rely on high costs of resistance, high
dominance of these costs and low dominance of the selection for the
resistance itself. These parameters are all not easy to measure and
there isn't agreement on their likely values {[}7{]}.

Note however that this combination of recessive resistance and high
dominance costs are exactly the features most desirable in any
insecticide irrespective of how it is deployed. We discuss fitness below
but note here that it is very difficult to maintain insecticide
concentration at the level required to make resistance recessive.
Concentrations decay post-application in most deployment strategies and
this cause resistance to gradually change from being recessive to being
dominant; see Figure 1 of {[}5{]} and our more extensive discussion in
{[}cite South et al WoS manuscript{]}.

The importance of resistance costs for substantial evolutionary benefits
of rotations over sequences points to the importance of the debate over
the fitness costs of insecticide resistance e.g. {[}7{]}. \emph{extend
this discussion of fitness costs}

We also found frequency of rotations had little effect. This is
important in the control of vectors of human diseases such as malaria
and dengue. These predominantly occur in resource-poor areas often with
poor healthcare infrastructure. Experience suggests that rotation
targets would often be missed (e.g rotations occur approximately every
15 generations rather than the planned ``official'' 10 generations) but
our simulations suggest this would not fatally undermine a rotations
policy.

\subsubsection{Caveats}\label{caveats}

The model does not include the potential effects of modifier genes
ameliorating the costs of resistance alleles. To include this would
require a more detailed model tracking linkage disequilibrium between
each resistance locus and its modifier. It has been suggested that
rotations could be favoured as a strategy because they keep resistance
frequences lower and restrict selection pressure for modifier genes
{[}3{]}. However, there is little evidence to support the existence of
such modifier genes in insectide resistance {[}{]}. Even with the
existence of modifier genes it is unlikely that they would provide much
of an advantage to rotations in the scenarios we describe. In the
modelled scenarios the use of an insecticide is stopped when the
resistance frequency for that insecticide reaches 0.5. Thus in all
scenarios resistance frequencies do not remain at high levels for long.
In this situation it seems unlikely that modifiers would be selected
for.

\subsubsection{Conclusions}\label{conclusions}

In their effect on the evolution of insecticide resistance, rotations
and sequences are most often the same. If costs of resistance are
included, infrequently (less than 7\% of model runs set between
plausible limits) there are substantial benefits to the rotation
strategy over a sequence. There was never a substantial benefit of a
sequence over a rotation. Whether or not insecticide rotations are
predicted to offer sizeable benefits over sequences is dependent on
coarse and fine scale issues about the nature of insecticide resistance
that are yet to be agreed upon. The model shows that resistance costs
are necessary to provide a substantial evolutionary advantage to
rotations. However even when these requirements are satisfied particular
combinations of inputs are required to generate the advantage of
rotations.

@@

\subsection{TODO}\label{todo}

\emph{add to methods description of how fitness determined in model,
move \& modify fig 1} \emph{try to come up with brief name for model
output to make it easier to refer to it later} \emph{add caveat about
polygenic resistance} \emph{The caption to each graph needs to have the
list of parameter values used to generate them} \emph{work on Ians fig Y
to help explanation}

@@

\subsection{Figures}\label{figures}

\begin{figure}

\includegraphics{pap3figs/Fig1-1} \hfill{}

\caption{Figure 1. [todo need to modify caption from MJ paper, start by saying that only in left panel during application of the insecticide] The effect of model inputs on the fitness of genotypes for a single insecticide. Fitness is shown on the y-axis and the different genotypes (SS, SR, RR) on the x axis. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS, resistance restoration 'restores' a portion of the fitness for RR and dominance of resistance determines how the fitness for SR lies between that of SS and RR. For those that are not exposed, fitness of SS is set to 1 by definition, resistance cost determines the fitness of RR and again dominance of cost determines how the fitness for SR sits between that of SS and RR. In this example effectiveness=0.8, resistance restoration=0.5 which 'restores' half of the fitness lost due to the insecticide, dominance of resistance=0.7 which sets the fitness of the SR closer to RR than SS. Resistance cost=0.3 which reduces fitness in the absence of the insecticide from 1 to 0.7, and dominance of cost=0.8 which sets fitness of SR close to RR.}\label{fig:Fig1}
\end{figure}
